---
title: "DEG + HMDB workflow"
author: "MetaNetAssoc Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DEG + HMDB workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Purpose

Version 0.2.0 starts with a differential-gene (DEG) set and differential
metabolites identified by HMDB IDs. The workflow is:

1. create a STRING PPI network from DEG;
2. filter an HMDB--gene MPI evidence table by DEG and metabolites;
3. perform DEG enrichment with `clusterProfiler`, exporting its result table;
4. score metabolite--term associations with `run_metanet()`; and
5. draw ranking, heatmap, and network figures.

`species` must be either `"human"` or `"mouse"` in all network-building
steps. The PPI source is STRING. MPI evidence must come from a traceable
metabolite--gene source; this package deliberately does not invent MPI edges.

## Input tables

```{r inputs, eval=FALSE}
# Either a vector ...
deg <- c("CHKA", "SLC44A1", "CPT1A")
hmdb <- c("HMDB0000122", "HMDB0000062")

# ... or tables with explicit columns.
deg_table <- data.frame(gene = deg, log2FC = c(1.2, -0.8, 1.1), padj = c(.01, .03, .02))
metabolite_table <- data.frame(hmdb_id = hmdb)

# Required MPI evidence-table columns (other columns are preserved as provenance).
my_mpi_database <- data.frame(
  hmdb_id = c("HMDB0000122", "HMDB0000062"),
  metabolite_name = c("Choline", "L-Carnitine"),
  relation_gene = c("CHKA", "CPT1A"),
  mpi_source = c("curated_source", "curated_source"),
  mpi_weight = c(.95, .90)
)
```

For a biological analysis, `my_mpi_database` must record the original source
and identifier mapping. The package's bundled MPI reference is synthetic and
is only for testing the code path.

## 1. Build PPI from STRING

With internet access, query STRING directly. STRING combined scores are on the
0--1000 scale; `build_ppi_network()` returns the standardized 0--1 score.

```{r string-online, eval=FALSE}
ppi <- build_ppi_network(
  deg_table,
  species = "human",
  gene_column = "gene",
  score_threshold = 400
)
```

For reproducibility, download and retain the STRING edge table, then supply it
with `string_data`. This skips all network access.

```{r string-offline, eval=FALSE}
ppi <- build_ppi_network(
  deg,
  species = "human",
  string_data = downloaded_string_edges,
  score_threshold = 400
)
```

## 2. Build the MPI network

```{r mpi, eval=FALSE}
mpi <- build_mpi_network(
  deg_table,
  metabolite_table,
  species = "human",
  mpi_database = my_mpi_database,
  gene_column = "gene",
  hmdb_column = "hmdb_id"
)
```

The output contains `metabolite_id`, `relation_gene`, `mpi_source`, and
`mpi_weight`, ready for `run_metanet()`.

## 3. Enrich DEG and export the source table

`run_deg_enrichment()` uses `clusterProfiler::enrichGO`. It needs
`clusterProfiler` plus `org.Hs.eg.db` for human or `org.Mm.eg.db` for mouse.
The `raw` component preserves the familiar `clusterProfiler` result table;
`enrichment` is the long term--gene representation used by this package.

```{r enrichment, eval=FALSE}
enr <- run_deg_enrichment(
  deg_table,
  species = "human",
  gene_column = "gene",
  ontology = "BP",
  export_file = "GO_enrichment_human.csv"
)

head(enr$raw)
```

## 4. Run the network scoring

```{r analysis, eval=FALSE}
fit <- run_metanet(
  metabolites = hmdb,
  mpi = mpi,
  ppi = ppi,
  enrichment = enr$enrichment,
  focus_genes = deg
)

head(result_scores(fit))
result_qc(fit)
```

If `deg` is provided, `run_metanet()` can construct any missing input. For a
reproducible analysis, passing the saved PPI and MPI tables explicitly is
recommended.

```{r one-call, eval=FALSE}
fit <- run_metanet(
  metabolites = hmdb,
  deg = deg,
  species = "human",
  mpi_database = my_mpi_database,
  string_data = downloaded_string_edges
)
```

## 5. Plot and inspect results

```{r plots, eval=FALSE}
plot_term_ranking(fit, term_id = result_scores(fit)$term_id[1])
plot_score_heatmap(fit)
plot_association_network(fit)

# Every supporting metabolite--gene--gene--term path is retained.
head(result_edges(fit))
```

## Human and mouse MPI files built with this project

The project-level files `MPI_human_KEGG_Reactome_HumanGEM_BRENDA.csv` and
`MPI_mouse_Ensembl_orthologs.csv` use KEGG compound identifiers rather than
HMDB IDs. Before using them with this HMDB-first workflow, add a documented
HMDB--KEGG cross-reference and expose the HMDB ID as `hmdb_id` (or rename it
to `metabolite_id`). The mouse table is an Ensembl orthology projection of
human evidence and should not be treated as independently curated mouse MPI
evidence.

In version 0.3.0 these references are also bundled with the package:

```{r packaged-reference, eval=FALSE}
human_mpi_kegg <- load_mpi_reference("human")
mouse_mpi_kegg <- load_mpi_reference("mouse")
```
