---
title: "MetaNetAssoc: from metabolites to biological terms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MetaNetAssoc: from metabolites to biological terms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Question answered by the package

MetaNetAssoc ranks candidate metabolites for a biological term by asking whether
genes associated with each metabolite are connected to member genes of that term
in a PPI network. The output is an association hypothesis with an auditable path,
not evidence of causal regulation.

## End-to-end example

```{r}
library(MetaNetAssoc)
example <- load_example_data()

fit <- run_metanet(
  metabolites = example$metabolites$metabolite_id,
  focus_genes = example$focus_genes,
  mpi = example$mpi,
  ppi = example$ppi,
  enrichment = example$enrichment,
  include_overlap = TRUE,
  hub_mode = "none"
)

fit
head(result_scores(fit))
result_qc(fit)
```

The example contains realistic pathway structure but synthetic evidence. It is
designed so that carnitine-related metabolites connect preferentially with energy
metabolism, glutathione and taurine with oxidative stress, and spermidine with
spermatid differentiation.

## Inspecting one association

```{r}
subset(
  result_edges(fit),
  metabolite_name == "Glutathione" & term_id == "GO:SIM0003"
)
```

This edge table is the primary audit trail. It shows the MPI gene, term gene,
evidence source, PPI confidence, path type, optional hub factor, and final edge
weight.

## Term size and hub sensitivity

`edge_count` is easy to understand but grows with the number of genes in either
set. `normalized_score` divides the weighted sum by the theoretical number of
gene pairs. The package reports both rather than hiding this trade-off.

```{r}
fit_penalized <- run_metanet(
  example$metabolites$metabolite_id,
  mpi = example$mpi,
  ppi = example$ppi,
  enrichment = example$enrichment,
  hub_mode = "penalize"
)
head(result_scores(fit_penalized))
```

## Optional term reduction

Terms can be reduced using Jaccard similarity of member genes. The most
significant term is retained as the representative.

```{r}
fit_reduced <- run_metanet(
  example$metabolites$metabolite_id,
  mpi = example$mpi,
  ppi = example$ppi,
  enrichment = example$enrichment,
  term_similarity_cutoff = 0.7
)
result_term_reduction(fit_reduced)
```

## Visual summaries

```{r, eval=requireNamespace("ggplot2", quietly=TRUE)}
plot_term_ranking(fit, "Response to oxidative stress")
plot_score_heatmap(fit)
```

